Rhizaria in the oligotrophic ocean exhibit clear temporal and vertical variability.

### 2024-06-26 manuscript/01_intro.qmd
edit manuscript

Rhizaria are an extremely diverse super-group of single-celled organisms consisting of several phyla including Retaria (foraminifera and radiolaria) and Cercozoa. These organisms exist in a wide range of habitats and are widely represented in plankton communities throughout the global ocean. While the taxonomy of these organisms has recently undergone several reclassifications [@biard2022], their presence in ocean ecosystems has been long known to oceanographers. Some of the earliest records of their existence are from oceanographic expeditions in the 19th century [@haekel1887]. Rhizaria are unique members of the plankton and protist community because they can reach large sizes (up to several mm in diameter) and they construct intricate mineral skeletons out of either silica, strontium, or calcium carbonate [@kimoto2015; @suzuki2015; @nakamura2015; @biard2022]. Despite their noticeable morphology and global distribution, Rhizaria were largely understudied throughout the 20th century. The bulk of modern plankton research has focused on hard-bodied crustacea which are numerically dominant and easily sampled with nets and preservatives. Fragile organisms like Rhizaria were difficult to adequately study as they can be destroyed through standard zooplankton sampling techniques and preserve poorly. A number of studies in the late 1900s did employ alternative techniques to quantify Rhizaria including diaphragm pumps [@michaels1988] or blue-water SCUBA collections [@caron1984; @caron1995; @bijma1990]. However, the bulk of Rhizaria research was constrained to sediment traps or paleontological studies of sediment [@takahashi1983; @boltovskoy1993]. Only recently has the advent of molecular techniques and in-situ imaging tools ignited a renewed focus on Rhizaria in pelagic ecosystems [@caron2016].

The wave of new data on Rhizaria has facilitated an improved understanding of the significance in ocean ecosystem functions. Firstly, taxonomists have been able to greatly refine the understanding of evolutionary relationships amongst these diverse protists [@aurahs2009; @decelle2012; @decelle2013; @biard2015; @cavalier-smith2018; rev by @biard2022]. DNA metabarcoding studies have revealed insights into the distributional patterns [@decelle2013; @biard2017; @marsbrisbin2020; @blanco-bercial2022; @llopismonferrer2022; @sogawa2022], ecological relationships [@decelle2012; @nakamura2023], and contribution to biogeochemical fluxes [@guidi2016; @gutierrez-rodriguez2019]. Transcriptomic and proteomic approaches also have been used to quantify Rhizaria contribution to community metabolism [@cohen2023]. Yet, despite the excellent taxonomic resolution provided by molecular approaches, they do not provide a truly quantitative metric for estimating Rhizaria abundance or biomass. In-situ imaging tools however, offer the ability to observe organisms in the natural state and quantify their abundance [@ohman2019a; @barth-inpress]. @biard2016 utilized in-situ imaging at a global scale to suggest Rhizaria were substantial contributors to the ocean carbon standing stock. While more recent calculations suggest lower carbon contribution [@laget2024], Rhizaria nonetheless have substantial influences on biogeochemical cycling. Due to their large sizes, ability to concentrate smaller particles and the unique structure of their mineral skeletons, Rhizaria have the potential to massively influence ocean biogeochemical cycling. A number of studies have made large advances in estimating the contribution of Rhizaria to ocean cycling of carbon [@lampitt2009; @stukel2018; @ikenoue2019; @gutierrez-rodriguez2019], silica [@biard2018; @llopismonferrer2021], and strontium [@decelle2013]. Still, Rhizaria ecological roles are not well understood [@biard2022]. This is a major challenge as it is critical to understand the ecological role of plankton to fully incorporate them into biological oceanographic models.

The ecological role of Rhizaria in plankton communities is complicated due to the fact different taxa can exhibit every different trophic modes. As zooplankton, rhizaria are predominately heterotrophic [@biard2022], yet their feeding modes can be quite varied. Phaeodarias (family Cercozoa) are largely thought to be flux-feeders, collecting and feeding on sinking particles [@nakamura2015; @stukel2019]. Alternatively, Retaria can be either exclusively heterotrophic or mixotrophic, utilizing photosynthetic algal symbionts [@anderson2014; @decelle2015]. Mixotrophic foraminifera host a variety of endosymbiont partners [@lee2006; @decelle2015], which are thought to support early and adult life stages and significantly contribute to total primary productivity [@kimoto2015]. Still, foraminifera are omnivorous, possibly even predominately carnivorous, with several studies suggesting that they can be effective predators [@anderson1976; @gaskell2019], mainly consuming live copepods [@caron1984]. Radiolaria have several lineages, all of which have some taxa well known to host symbionts [@biard2022b]. Amongst Radiolarias, arguably the most widespread are Collodaria who can be either large solitary cells or form massive colonies, up to several meters in length [@swanberg1981]. All known Collodaria species host dinoflagellate symbionts [@biard2022b] and can contribute substantially to primary productivity, particularly in oligotrophic ocean regions [@caron1995; @dennett2002]. This Collodaria-symbiont association has been suggested as a reason for their high abundances throughout the photic zone of oligotrophic environments globally [@biard2016; @biard2017]. A few Acantharea (Radiolaria order) clades host algal symbionts [@decelle2012; @biard2022b], notably with two clades forming an exclusive relationship with *Phaeocystis*. However, globally, Acantharea are less abundant than Collodaria [@biard2022] and contribute less to total primary productivity [@michaels1995]. This may be due to the fact several clades of Acantharea are cyst-forming and strictly heterotrophs [@decelle2013; @biard2022b]. Furthermore, @marsbrisbin2020 documented apparent predation behavior in Acantharea near the surface, suggesting that there may be a larger reliance on carnivory.

Given the high abundances, yet diverse trophic strategies found among Rhizaria taxa, it is reasonable to expect some form of niche partitioning. A number of studies do suggest evidence for vertical zonation between Rhizaria groups according to various trophic strategies. Taxa-specific studies of Radiolarias suggest they may be restricted to the euphotic zone [@michaels1988; @boltovskoy2017]. Although some studies report Acantharea in deeper waters [@decelle2013; @gutiérrez-rodríguez2022], Phaeodarias, alternatively, are generally found in the mesopelagic where photosynthesis cannot occur, but they can feed on sinking particles [@stukel2018]. In an imaging-based study of the whole Rhizaria community, @biard2020 noted clear patterns in vertical zonation which largely corresponded to different trophic roles. In the oligotrophic ocean, @blanco-bercial2022 also noted that the protist community, including Rhizaria, partition along an autotroph and mixotroph to heterotroph gradient with increasing depth in the water column. Yet, few studies have made direct attempts to relate rhizaria abundances to environmental factors [@biard2020]. In part, this is due to the fact few studies have been able to sample Rhizaria in the same location over a consistent timeframe [@michaels1988; @boltovskoy1993; @michaels1995; @hull2011; @gutiérrez-rodríguez2022]. Furthermore, no studies have utilized imaging, arguably the best method for quantifying rhizaria, consistently throughout the full mesopelagic. Given this lack of information, there are many unknowns with respect to Rhizaria ecology, seasonality and phenology across different groups.

In this study, we present a comprehensive assessment of large Rhizaria measured for over a year from regularly occurring cruises at monthly intervals. We utilized an in-situ imaging approach to facilitate abundance calculations. With this dataset, we address two critical aims. 1) Quantification of large Rhizaria throughout the epipelagic (0-200m) and mesopelagic (200-1000m) over the course of an annual cycle. These data were collected in the Sargasso Sea, and represents the first study of its kind in an oligotrophic system; and 2) We aim to test the hypothesis that Rhizaria exhibit niche partitioning according to trophic roles. This hypothesis makes several predictions, including vertical zonation, as seen in prior studies, but also that environmental variables related to trophic strategy will explain abundance patterns. Specifically, autotrophic/mixotrophic taxa will correspond to variables related to autotrophy (chl-a concentration, primary productivity, local DO maxima) and other rhizaria will correspond to factors which promote heterotrophy (particle concentration, flux, and local DO minima).

>>>>

Rhizaria are an extremely diverse super-group of single-celled organisms which occupy a wide range of habitats. Planktonic Rhizaria include Foraminifera, Radiolaria (including Acantharea and Collodaria) and Phaeodaria, all of which are observed throughout the global ocean [@biard2016; @laget2024] . While the taxonomy of these organisms has recently undergone several reclassifications [@biard2022], their presence in ocean ecosystems has been long known to oceanographers. Some of the earliest records of their existence are from oceanographic expeditions in the 19th century [@haekel1887]. Rhizaria are unique members of the plankton and protist community because they can reach large sizes (up to several mm in diameter) and they construct intricate mineral skeletons out of either silica, strontium, or calcium carbonate [@kimoto2015; @suzuki2015; @nakamura2015; @biard2022]. Despite their noticeable morphology and global distribution, Rhizaria were largely understudied throughout the 20th century. Fragile organisms like Rhizaria are difficult to adequately study because they can be destroyed through standard zooplankton sampling techniques and preserve poorly. A number of studies in the late 1900s did employ alternative techniques to quantify Rhizaria including diaphragm pumps [@michaels1988] or blue-water SCUBA collections [@caron1984; @bijma1990; @caron1995]. However, the bulk of Rhizaria research was constrained to sediment traps or paleontological studies of sediment [@takahashi1983; @boltovskoy1993].

Recently, the advent of molecular techniques and *in situ* imaging tools have ignited a renewed focus on Rhizaria in pelagic ecosystems [@caron2016]. Molecular tools have shed light on Rhizaria taxonomy [@biard2022], distribution [@blanco-bercial2020; @sogawa2022], ecology [@decelle2012; @nakamura2023], biogeochemical roles [@gutierrez-rodriguez2019; @laget2024], and metabolism [@cohen2023]. Yet, despite the excellent taxonomic resolution provided by molecular approaches, they do not provide a truly quantitative metric for estimating Rhizaria abundance or biomass. *In situ* imaging tools however, offer the ability to observe organisms in the natural state and quantify their abundance [@ohman2019a; @barth2024]. @biard2016 utilized *in situ* imaging at a global scale to suggest large (\>500$\mu$m diameter) Rhizaria were substantial contributors to the ocean carbon standing stock. While more recent calculations suggest lower carbon contribution [@laget2024], Rhizaria nonetheless have substantial influences on biogeochemical cycling. This influence is due to their unique mineral skeletons and ability to repackage small particles into larger, fast sinking ones [@ikenoue2019]. Thus, some Rhizaria may play substantial roles in ocean export through their interception of sinking particles [@stukel2018; @laget2024]. However, Rhizaria ecology is poorly understood [@biard2022] and their ecological roles are likely more than diverse than simple particle interception.

The ecological role of Rhizaria in plankton communities is complicated due to the fact different taxa can exhibit very different trophic modes. Many Rhizaria are strictly heterotrophic [@biard2022], yet their feeding modes can be quite varied. Phaeodaria are largely thought to be flux-feeders, collecting and feeding on sinking particles [@gowing1989]. Alternatively, Retaria can be either exclusively heterotrophic or mixotrophic, utilizing photosynthetic algal symbionts [@anderson2014; @decelle2015]. Mixotrophic Foraminifera host a variety of endosymbiont partners [@lee2006; @decelle2015], which are thought to support early and adult life stages and significantly contribute to total primary productivity [@kimoto2015]. Still, Foraminifera are omnivorous, possibly even predominately carnivorous, with several studies suggesting that they can be effective predators [@anderson1976; @gaskell2019], mainly consuming live copepods [@caron1984]. Radiolaria have several lineages, many of which host symbionts [@biard2022b]. Amongst Radiolaria, arguably the most widespread are Collodaria who can be either large solitary cells or form massive colonies, up to several meters in length [@swanberg1981]. All known Collodaria species host dinoflagellate symbionts [@biard2022b] and can contribute substantially to primary productivity, particularly in oligotrophic ocean regions [@caron1995; @dennett2002]. This Collodaria-symbiont association has been suggested as a reason for their high abundances throughout the photic zone of oligotrophic environments globally [@biard2016; @biard2017]. A few Acantharea (Radiolaria order) clades host algal symbionts [@decelle2012; @biard2022b], notably with two clades forming an exclusive relationship with *Phaeocystis*. However, globally, Acantharea are less abundant than Collodaria [@biard2022] and contribute less to total primary productivity [@michaels1995]. This may be due to the fact several clades of Acantharea are cyst-forming and strictly heterotrophs [@decelle2013; @biard2022b]. Furthermore, @marsbrisbin2020 documented apparent predation behavior in Acantharea near the surface, suggesting that there may be a larger reliance on carnivory.

Given the high abundances, yet diverse trophic strategies found among Rhizaria taxa, it is reasonable to expect some form of niche partitioning. A number of studies do suggest evidence for vertical zonation between Rhizaria groups according to various trophic strategies. Taxon-specific studies of Radiolaria suggest they may be restricted to the euphotic zone [@michaels1988; @boltovskoy2017]. Although some studies report Acantharea in deeper waters [@decelle2013; @gutiérrez-rodríguez2022], Phaeodaria, alternatively, are generally found beyond depths where light can sustain photosynthesis, but they can feed on sinking particles [@stukel2018; @laget2024]. In an imaging-based study of the whole Rhizaria community, @biard2020 noted clear patterns in vertical zonation which largely corresponded to different trophic roles. Protists communities, including Rhizaria, have been described to partition along autotroph-mixotroph-heterotroph gradients with increasing depth in the water column [@blanco-bercial2022; @laget2024]. Yet, few studies have made direct attempts to relate Rhizaria abundances to environmental factors [@biard2020]. In part, this is due to the fact few studies have been able to sample Rhizaria in the same location over a consistent timeframe [@michaels1988; @boltovskoy1993; @michaels1995; @hull2011; @gutiérrez-rodríguez2022]. Furthermore, very few studies have utilized *in situ* imaging, arguably the best method for quantifying Rhizaria, to consistently throughout the full mesopelagic [@laget2024] . Given this lack of information, there are many unknowns with respect to Rhizaria ecology, seasonality and phenology across different groups.

In this study, we present a comprehensive assessment of large Rhizaria abundance measured for over a year from regularly occurring cruises at monthly intervals. We utilized an *in situ* imaging approach to facilitate abundance calculations. With this dataset, we address two critical aims. 1) Quantifying of large Rhizaria throughout the epipelagic (0-200m) and mesopelagic (200-1000m) over the course of an annual cycle. These data were collected in the Sargasso Sea, and represents the first study of its kind in an oligotrophic system; and 2) We aim to test the hypothesis that Rhizaria exhibit niche partitioning according to trophic roles. This hypothesis has several implications, including vertical zonation, as seen in prior studies, but also that environmental variables related to trophic strategy will explain abundance patterns. Specifically, autotrophic/mixotrophic taxa will correspond to variables related to autotrophy (chl-a concentration, primary productivity, local DO maxima) and other rhizaria will correspond to factors which promote heterotrophy (particle concentration, flux, and local DO minima).

### 2024-06-26 manuscript/02_methods.qmd
edit manuscript

## Oceanographic Sampling

Data were collected in collaboration with the Bermuda Atlantic Time-series Study [@michaels1996; @lomas2013] on board the R/V Atlantic Explorer. Cruises were conducted at approximately monthly intervals. Rhizaria individuals were sampled using the Underwater Vision Profiler 5 [UVP5; @picheral2010], a tool which is well established to accurately quantify large Rhizaria [@biard2016; @stukel2018; @stukel2019; @biard2020; @barth2022; @drago2022; @llopismonferrer2022; @panaïotis2023]. The UVP5 was mounted to the sampling rosette and collected data autonomously on routine casts, from which only the downcast data are utilized. The UVP5 was deployed from June-September 2019 then from October 2020 - January 2022, during which time the BATS region was sampled for 3-5 days at monthly intervals. Casts were filtered to only include data collected in the BATS region, far offshore of Bermuda in the Sargasso Sea (approximately 31.0$^oN$-32.5$^oN$, 64.25$^oW$-63$^oW$; Supplemental Figure 1). In general, casts extended to either 200m, 500m, or 1200m deep, with a few extended into the bathypelagic (4500m). However, Rhizaria were only typically found in large abundances throughout the epipelagic and mesopelagic zones. As such, we limit this study to results from the upper 1000m of the water column.

A variety of biotic and abiotic data were collected during each BATS cruise. Briefly, we will explain the data utilized in this study. The UVP5 provided particle count data at a high-frequency from each cast. Particle concentration was calculated from this data for all particles 184$\mu m$ - 450$\mu m$. The lower size range was set by what could be reliably sampled by the UVP5's pixel resolution (\>2px; 0.092mm per pixel) and the upper size range is representative of a potential prey field for mesozooplankton [@whitmore2021]. For each UVP cast supporting continuous profiles of the CTD parameters salinity, temperature, and auxiliary CTD channels; Dissolved Oxygen (DO), in-situ chlorophyll fluorescence were measured at 24Hz using the BATS CTD package. On select casts, Niskin bottles were used to collect bacterial abundance estimates (via epifluorescence microscopy) as well as measure inorganic nutrients ($NO_3$, and $Si$ as silicate/sicilic acid) at discrete depths. On each cruise, flux estimates of total mass, organic carbon, and nitrogen were also collected using sediment traps; in the present study we utilized flux to the mesopelagic as the flux at 200m. Also primary productivity was estimated through measuring $^{14}C$ uptake rates from in-situ incubations. For full descriptions of the BATS sampling program and methods, see @knap1997 and @lomas2013 for a review. Additionally, data can be viewed online [(https://bats.bios.asu.edu/bats-data/)](https://bats.bios.asu.edu/bats-data/).

Environmental data were processed in a variety of ways to match the format of the Rhizaria abundance estimates (see below). CTD data were collected at higher frequency than the UVP (24Hz vs 15Hz respectively), so these data were averaged within matching bins to the UVP5 data. Data from Niskin bottles were first linearly interpolated in depth at 1m resolution then time averaged over the cruise, then subsequently averaged into matching UVP5-sized bins. Primary productivity estimates were totaled within the euphotic zone to represent a "total euphotic productivity".

## Rhizaria imaging processing and quantification

Individual vignettes of Rhizaria images were identified using the classification platform Ecotaxa [@picheral]. Data were pre-sorted utilizing a random-forest classifier and pre-trained learning set. Taxonomic classification were done based on morphology exclusively. While there are sparse taxonomic guides for in-situ images of rhizaria, identification largely relied on descriptions in [@nakamura2015; @suzuki2015; and @biard2020]. Using the aforementioned sources and publicly available ecotaxa projects, we constructed a guide accessible at: <https://thealexbarth.github.io/media//Project_Items/Oligotrophic_Community/ecotaxa_UVP-guide-stone-lab.pdf>. Broadly, Rhizaria were classified as Foraminifera, Radiolarias (Acantharea or Collodaria), or as a variety of Phaeodaria families (Figure 1). When identification could not be confidently made between a few candidate taxa, a less specific label was used. As a result, we have data from "unidentified Rhizaria", which typically were vignettes not distinguishable between Aulacanthidae or Acantharea or "unidentified Phaeodaria", which are clearly Phaeodaria but not distinguishable into a family.

![Example images of different Rhizaria taxa. 4mm scale bar shown in lower right. All vignettes are same scale](images/01_taxa.pdf)

The UVP5 samples at \~15Hz rate as it descends the water column and records the exact position of each particle larger than 600$\mu m$. However, identified rhizaria ranged from a 934$\mu m$ Aulacanthidae cell to a Collodarian colony over 10mm in diameter. To confirm that the UVP5 was sampling adequately across all size ranges, an normalized biomass size spectrum (NBSS) slope was constructed to identify a drop-off which would indicate poor-sampling at the small size range [@lombard2019; @barth-inpress]. However, it was evident from this analysis that all size ranges were adequately sampled across the size range (Supplemental Figure 2) so no data were excluded. The UVP5 reports the exact depth at which a particle is recorded, however to estimate abundance, observations must be binned over fixed depth intervals. Our deployments had variable descent depths and speeds with more casts descending to 500m than 1000m and descents quicker through the epipelagic than the mesopelagic (see @barth2022 for an extended discussion of UVP5 data processing). For the present study, Rhizaria abundances were estimated in 25m vertical bins, which offer a moderate sampling volume per bin (average 0.948$m^3$ in the epipelagic and 0.589$m^3$ in the mesopelagic; Supplemental Table 1) while still maintaining ecologically relevant widths. However, concentrations in a 25m bin would need to be greater than 2.428 ind. $m^{-3}$ and 3.912 ind. $m^{-3}$, in the epipelagic and mesopelagic respectively, to fall below a 10% non-detection risk [@benfield1996; @barth-inpress]. Because we typically observed many rhizaria taxa below these concentrations, we present the 25m binned data to visualize broad-scale average distributions. For quantifying and modelling Rhizaria abundances, we present integrated abundance estimates, with each cast. Due to the variable descent depths of the UVP, data are categorized as epipelagic (0-200m), upper mesopelagic (200-500m), and lower mesopelagic (500-1000m). The average sampling volume integrated through these regions were 7.59$m^2$, 7.06$m^2$, and 11.77$m^2$, with non-detection thresholds at 0.30 ind. $m^{-2}$, 0.33 ind. $m^{-2}$, and 0.20 ind. $m^{-2}$ respectively. All UVP data processing was done using the `EcotaxaTools` package in R (Barth 2023).

## Modelling environmental controls of Rhizaria Abundance

Generalized Additive Models (GAMs) were used to assess the relationship between integrated Rhizaria abundance and different environmental factors. GAMs offer the ability to model non-linear and non-monotonic relationships, which can be particularly useful in assessing ecological relationships [@wood2017] and have been successfully applied to Rhizaria ecology [@biard2020]. The `mgcv` package [@wood2001] was used to construct models relating environmental parameters to each taxonomic group's integrated abundance estimates from each cast. To select the most parsimonious model for each analysis, a backwards step-wise approach was taken. First, a full model was fit using any term which may be ecologically relevant. Terms were fit using maximum likelihood with a double penalty approach on unnecessary smooths [@marra2011]. The smoothness parameter was restricted (k = 6) to prevent overfitting the models. At each iteration of the backwards step-wise procedure, the model term with the lowest F score (least statistically significant) was removed. This was repeated until all model terms were statistically significant or the $R^2_{adj}$ was substantially reduced. Models were fit for each region; epipelagic, upper mesopelagic, and lower mesopelagic. In cases where observations were too sparse for a given taxonomic grouping, models were not run. All code and full models are available in code, as well as intermediate data products at <https://github.com/TheAlexBarth/RhizariaSeasonality>.

>>>>

## Oceanographic sampling

Data were collected in collaboration with the Bermuda Atlantic Time-series Study [@michaels1996; @lomas2013] on board the R/V Atlantic Explorer. Cruises were conducted at approximately monthly intervals. Rhizaria individuals were sampled using the Underwater Vision Profiler 5 (UVP) [@picheral2010]. The UVP5 is an in-situ camera used to capture plankton and particle images and has been well established to accurately quantify large Rhizaria abundance [@biard2016; @stukel2018; @stukel2019; @biard2020; @barth2022; @drago2022; @llopismonferrer2022; @panaïotis2023]. The UVP5 was mounted to the sampling rosette and collected data autonomously on routine casts, from which only the downcast data are utilized. The UVP5 was deployed from June-September 2019 then from October 2020 - January 2022, during which time the BATS region was sampled for 3-5 days at monthly intervals (see sampling details in Supplemental Table 1). Casts were filtered to only include data collected in the BATS region, far offshore of Bermuda in the Sargasso Sea (approximately 31.0$^oN$-32.5$^oN$, 64.25$^oW$-63$^oW$; Supplemental Figure 1). In general, casts extended to either 200m, 500m, or 1200m deep, with a few extended into the bathypelagic (4500m). However, Rhizaria were only typically found in large abundances throughout the epipelagic and mesopelagic zones. As such, we limit this study to results from the upper 1000m of the water column.

A variety of biotic and abiotic data were collected during each BATS cruise. The UVP5 provided particle count data at a high-frequency from each cast. Particle concentration, as a proxy for prey field [@whitmore2021], was calculated from this data for all particles 184$\mu m$ - 900$\mu m$. The lower size range was set by what could be reliably sampled by the UVP5's pixel resolution (\>2px; 0.092mm per pixel) and the upper size range was set to below the size of observable Rhizaria yet large enough to be inclusive of particles on which some Rhizaria may feed. For each UVP cast supporting continuous profiles of the CTD parameters salinity, temperature, and auxiliary CTD channels; Dissolved Oxygen (DO), *in situ* chlorophyll fluorescence were measured at 24Hz using the BATS CTD package. On select casts, Niskin bottles were used to collect bacterial abundance estimates (via epifluorescence microscopy) as well as measure inorganic nutrients ($NO_3$, and $Si$ as silicate/sicilic acid) at discrete depths. On each cruise, flux estimates of total mass, organic carbon, and nitrogen were also collected using sediment traps; in the present study we utilized flux to the mesopelagic as the flux at 200m. Also primary productivity was estimated through measuring $^{14}C$ uptake rates from *in situ* incubations. For full descriptions of the BATS sampling program and methods, see @knap1997 and @lomas2013 for a review. Additionally, data can be viewed online [(https://bats.bios.asu.edu/bats-data/)](https://bats.bios.asu.edu/bats-data/).

Environmental data were processed to match the format of the Rhizaria abundance estimates (see below). CTD data were collected at higher frequency than the UVP (24Hz vs 15Hz respectively), so these data were averaged within matching bins to the UVP5 data. Data from Niskin bottles were first linearly interpolated in depth at 1m resolution then time averaged over the cruise, then subsequently averaged into matching bins as the UVP5 data. Primary productivity estimates were depth integrated throughout the euphotic zone (0-140m).

## Rhizaria imaging processing and abundance quantification

Individual vignettes of Rhizaria images were identified using the classification platform EcoTaxa [@picheral]. Data were pre-sorted utilizing a random-forest classifier and pre-trained learning set. Taxonomic classification were done based on morphological parameters measured in Zooprocess [@gorsky2010], which includes parameters such as major axis, equivalent spherical diameter (ESD), and grey values. While there are sparse taxonomic guides for *in situ* images of Rhizaria, identification largely relied on descriptions in [@nakamura2015; @suzuki2015; and @biard2020]. Using the aforementioned sources and publicly available EcoTaxa projects, we constructed a guide accessible at: <https://thealexbarth.github.io/media//Project_Items/Oligotrophic_Community/ecotaxa_UVP-guide-stone-lab.pdf>. Broadly, Rhizaria were classified as Foraminifera, Radiolaria (Acantharea or Collodaria), or as a variety of Phaeodaria families (Figure 1). When identification could not be confidently made between a few candidate taxa, a less specific label was used. As a result, we have data from "unidentified Rhizaria", which typically were vignettes not distinguishable between Aulacanthidae or Acantharea or "unidentified Phaeodaria", which are clearly Phaeodaria but not distinguishable into a family.

![Example images of different Rhizaria taxa as collected by UVP5 imaging. 4mm scale bar shown in lower right. All vignettes are same scale](images/01_taxa.pdf)

The UVP5 samples at \~15Hz rate as it descends the water column and records the exact position of each particle larger than 600$\mu m$ (ESD). However, identified rhizaria ranged from a 934$\mu m$ ESD Aulacanthidae cell to a Collodarian colony over 10mm in diameter. To confirm that the UVP5 was sampling adequately across all size ranges, an normalized biomass size spectrum (NBSS) slope was constructed to identify a drop-off which would indicate poor-sampling at the small size range [@lombard2019; @barth2024]. However, it was evident from this analysis that all size ranges were adequately sampled across the size range (Supplemental Figure 2) so no data were excluded. The UVP5 reports the exact depth at which a particle is recorded, however to estimate abundance, observations must be binned over fixed depth intervals. Our deployments had variable descent depths and speeds with more casts descending to 500m than 1000m and descents quicker through the epipelagic than the mesopelagic (see @barth2022 for an extended discussion of UVP5 data processing). For the present study, Rhizaria abundances were estimated in 25m vertical bins, which offer a moderate sampling volume per bin (average 0.948$m^3$ in the epipelagic and 0.589$m^3$ in the mesopelagic; Supplemental Table 1) while still maintaining ecologically relevant widths. However, concentrations in a 25m bin would need to be greater than 2.428 ind. $m^{-3}$ and 3.912 ind. $m^{-3}$, in the epipelagic and mesopelagic respectively, to fall below a 10% non-detection risk [@benfield1996; @barth2024]. Because we typically observed many Rhizaria taxa below these concentrations, we present the 25m binned data to visualize broad-scale average distributions. For quantifying and modelling Rhizaria abundances, we present integrated abundance estimates, with each cast. Due to the variable descent depths of the UVP, data are categorized as epipelagic (0-200m), upper mesopelagic (200-500m), and lower mesopelagic (500-1000m). The average sampling volume integrated through these regions were 7.59$m^2$, 7.06$m^2$, and 11.77$m^2$, with non-detection thresholds at 0.30 ind. $m^{-2}$, 0.33 ind. $m^{-2}$, and 0.20 ind. $m^{-2}$ respectively. All UVP data processing was done using the `EcotaxaTools` package in R (Barth 2023).

## Modelling environmental controls of Rhizaria abundance

Generalized Additive Models (GAMs) were used to assess the relationship between integrated Rhizaria abundance and different environmental factors. GAMs offer the ability to model non-linear and non-monotonic relationships, which can be particularly useful in assessing ecological relationships [@wood2017] and have been successfully applied to Rhizaria ecology [@biard2020]. The `mgcv` package [@wood2001] was used to construct models relating environmental parameters to each taxonomic group's integrated abundance estimates from each cast. To select the most parsimonious model for each analysis, a backwards step-wise approach was taken. First, a full model was fit using any term which may be ecologically relevant. Terms were fit using maximum likelihood with a double penalty approach on unnecessary smooths [@marra2011]. The smoothness parameter was restricted (k = 6) to prevent over fitting the models. Then, the full model was checked for concurvity (a metric indicating high parameter relatedness in GAMs). If two terms had a high concurvity value (\>0.8), then two alternative models were constructed each leaving out one of the related terms. The more parsimonious model of these two was selected using AIC. Once concurvity of terms was reduced, then the model was subjected to the backwards step-wise procedure. At each iteration, the model term with the lowest F score (least statistically significant) was removed. This was repeated until all model terms were statistically significant or the $R^2_{adj}$ was substantially reduced. Models were fit for each water column region; epipelagic, upper mesopelagic, and lower mesopelagic. In cases where observations were too sparse for a given taxonomic grouping, models were not run. All code, full models, and model selection tools are available in open-source scripts, as well as intermediate data products at <https://github.com/TheAlexBarth/RhizariaSeasonality>.
